Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019

### 2020-05-14 manuscript/EP1162_manuscript_draft.Rmd
more manny work

# Load libraries
library(tidyverse)
library(here)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)
library(lsmeans)

@@

volPlot<-
  standDat %>% 
  # plot
  ggplot()+
  aes(x=Year,y=BA,fill=BA.Target)+
  geom_boxplot()+
  facet_wrap(~TU)+
  theme(legend.position="none")+
  ylab("Plot volume (m³/hectare)")

@@

Annual increment between the four periods differed between treatment units (`r figNums("incPlot",display="cite")`).  In the high-removal treatment unit, annual basal area increment increased in each subsequent measurement period, with the highest mean plot values of 0.9m^2^ per hectare.  In contrast, mean plot basal area annual increment was highest in the low-removal treatment unit during the 1994-1997 period, and decreased in subsequent periods.  A similar pattern was observed in the control treamtent unit.  Annual increment patterns in total and merchantable volumes were similar to basal area, with the exception of the high-removal treatment unit.  Here, maximum annual volume increment was highest between 1997 and 2009, with a mean plot value of 8m^3^ per year per hectare; this increment decreased slightly from 2009-2019 period.  In the control unit, volume increment from 2009-2019 was lower than from 1992-1994.

@@

```{r compare sph and BA differences between periods and diam classes}
perComp<-
  dat %>%
  filter(Year %in% c(1992,2019)) %>% # filter for start and end years
  group_by(Plot,Year) %>% 
  summarise(N.Plot=n()) %>% # summarise number of years per plot
  group_by(Plot) %>% 
  summarise(N.Plot=n()) %>% 
  filter(N.Plot==2) %>% 
  pull(Plot) %>% 
  data.frame(Plot=.) %>%

@@

  # break trees into diameter classes
  mutate(DBH.class=cut(DBH,breaks=c(7.5,12.5,18.5,30,90),include.lowest = TRUE,right=FALSE)) %>% 
  left_join(plotDat,by="Plot")

# fit models
fit1<-  
  perComp %>%

  # come up with a denominator (number of plots per BA.Target per year)
  distinct(Plot,.keep_all = TRUE) %>% 
  group_by(BA.Target) %>% 
  summarise(Num.Plot=n()) %>% 
  right_join(perComp,by="BA.Target") %>%

  # Plot averages
  group_by(Plot,Year,BA.Target,Num.Plot,Species,DBH.class) %>% 
  summarise(BA=sum(BA)*20,# Notice the plot multiplier
            Vol=sum(VolTot)*20,
            SPH=n()*20) %>% 
  filter(DBH.class=="[30,90]") %>% 
  filter(Species=="Sx") %>% 
  lm(BA~BA.Target+Year,data=.)

# plotting
perCompPlot<- # assign variable  
  perComp %>%

  # come up with a denominator (number of plots per BA.Target per year)
  distinct(Plot,.keep_all = TRUE) %>% 
  group_by(BA.Target) %>% 
  summarise(Num.Plot=n()) %>% 
  right_join(perComp,by="BA.Target") %>%

  # Plot averages
  group_by(Plot,Year,BA.Target,Num.Plot,Species,DBH.class) %>% 
  summarise(BA=sum(BA)*20,# Notice the plot multiplier
            Vol=sum(VolTot)*20,
            SPH=n()*20) %>%

  #BA Target averages
  group_by(BA.Target,Year,Species,Num.Plot,DBH.class) %>% 
  summarise(BA.sum=sum(BA),
            BA.se=sd(BA)/sqrt(length(BA)),
            Vol.sum=sum(Vol),
            Vol.se=sd(Vol)/sqrt(length(Vol)),
            SPH.sum=sum(SPH),
            SPH.se=sd(SPH)/sqrt(length(SPH))) %>% 
  ungroup() %>%

  # Select only for Sx and Bl
  filter(Species %in% c("Bl","Sx")) %>%

  # Compute means
  mutate(BA.mean=BA.sum/Num.Plot,
         Vol.mean=Vol.sum/Num.Plot,
         SPH.mean=SPH.sum/Num.Plot) %>%

  # format
  dplyr::select(BA.Target,Year,Species,DBH.class,BA=BA.mean,Vol=Vol.mean,SPH=SPH.mean) %>% 
  pivot_longer(cols=BA:SPH,names_to="var",values_to="value")

vol.plot2<-  
  perCompPlot %>% 
  filter(var=="Vol") %>%

  # Plotting
  ggplot()+
  aes(x=DBH.class,y=value,fill=Species,group=Species)+
  geom_bar(stat="Identity",position="dodge")+
  facet_grid(Year~BA.Target)

BA.plot2<-  
  perCompPlot %>% 
  filter(var=="BA") %>%

  # Plotting
  ggplot()+
  aes(x=DBH.class,y=value,fill=Species,group=Species)+
  geom_bar(stat="Identity",position="dodge")+
  facet_grid(Year~BA.Target)

SPH.plot2<-  
  perCompPlot %>% 
  filter(var=="SPH") %>%

  # Plotting
  ggplot()+
  aes(x=DBH.class,y=value,fill=Species,group=Species)+
  geom_bar(stat="Identity",position="dodge")+
  facet_grid(Year~BA.Target)

@@

`r figNums("volPlot.plot",caption="Variation in volume per plot (m³ per hectare) at five measurement years, faceted by treatment unit.")`

@@

>>>>

# Load libraries
library(tidyverse)
library(here)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)
library(lsmeans)
library(cowplot)

@@

volPlot<-
  standDat %>% 
  # plot
  ggplot()+
  aes(x=Year,y=Vol.Tot,fill=BA.Target)+
  geom_boxplot()+
  facet_wrap(~TU)+
  theme(legend.position="none")+
  ylab("Plot volume (m³/hectare)")

@@

Annual increment between the four periods differed between treatment units (`r figNums("incPlot",display="cite")`).  In the first three growing seasons after treatment, basal area and volume increments in the two treatment units with basal area removal were lower than increment in the control unit.  In the high-removal treatment unit, basal area increased in each subsequent measurement period, and exceeded the other units between 1997 and 2019.  In the low-removal and control treatment units, increment either declined or stabilized after ca.1997.  Volume increment declined in all units from 2009-2019, and in the control unit, annual volume increment was lower in 2019 than in 1992.

@@

```{r compare species proportions as per Deal 2002}

sppProp.Dat<-
  dat %>%
  filter(Year %in% c(1992,2019)) %>% # filter for start and end years
  group_by(Plot,Year) %>% 
  summarise(N.Plot=n()) %>% # summarise number of years per plot
  group_by(Plot) %>% 
  summarise(N.Plot=n()) %>% 
  filter(N.Plot==2) %>% 
  pull(Plot) %>% 
  data.frame(Plot=.) %>%

@@

  # break trees into diameter classes
  mutate(DBH.class=cut(DBH,breaks=c(7.5,12.5,18.5,30,90),include.lowest = TRUE,right=FALSE)) %>%

  # Calculate plot BA, Volume and SPH by species, plot and DBH.class 
  group_by(Plot,Species,Year,DBH.class) %>%
   summarise(BA=sum(BA)*20,# Notice the plot multiplier
            Vol=sum(VolTot)*20,
            SPH=n()*20) %>%

  # Calculate species proportions
    filter(Species %in% c("Bl","Sx")) %>% # filter for Sx and Bl
    pivot_longer(cols=BA:SPH,names_to="Var",values_to="Value") %>% 
 pivot_wider(names_from=Species,values_from=Value) %>% 
  replace_na(list(Bl=0,Sx=0))      # replace NA with 0

# Plotting
sppProp.Plot<-
  sppProp.Dat %>% 
  mutate(Total=Bl+Sx) %>% 
  mutate(Bl=Bl/Total,
         Sx=Sx/Total) %>% 
  pivot_longer(cols=Bl:Sx,names_to="Species",values_to="Prop") %>% 
  left_join(plotDat,by="Plot") %>% 
  group_by(BA.Target,Year,DBH.class,Var,Species) %>% 
  summarise(Prop=mean(Prop)) %>%
  mutate(TU=paste("TU",BA.Target,sep=":")) %>% 
  filter(Var=="Vol") %>% 
  ggplot()+
  aes(x=DBH.class,y=Prop,group=Species,fill=Species)+
  geom_bar(stat="Identity")+
  facet_grid(Year~TU)+
  ylab("Proportion")+
  xlab("Diameter class (cm)")+
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 270,vjust=0.2,hjust=0))

sppVol.Plot<-
  sppProp.Dat %>% 
  pivot_longer(cols=Bl:Sx,names_to="Species",values_to="Val") %>% 
  left_join(plotDat,by="Plot") %>% 
  group_by(BA.Target,Year,DBH.class,Var,Species) %>% 
  summarise(Val=mean(Val)) %>%
  mutate(TU=paste("TU",BA.Target,sep=":")) %>% 
  filter(Var=="Vol") %>% 
  ggplot()+
  aes(x=DBH.class,y=Val,group=Species,fill=Species)+
  geom_bar(stat="Identity")+
  facet_grid(Year~TU)+
  ylab("Volume (m³/ha)")+
  xlab("Diameter class (cm)")+
  theme(legend.position="none",
        axis.title.x=element_blank(),
        axis.text.x=element_blank())

sppProp.plot<-plot_grid(sppVol.Plot,sppProp.Plot,nrow=2,rel_heights=c(1,1.5),axis="lr",align="h")

figNums("sppProp.plot",display=FALSE)
```

@@

`r figNums("volPlot.plot",caption="Variation in total volume per plot (m³ per hectare) at five measurement years, faceted by treatment unit.")`

@@

##### page break
```{r plot total volume species proportion,fig.height=6}

sppProp.plot

`r figNums("sppProp.plot",caption="Proportion of mean plot volume by four diameter classes of fir and spruce, expressed as volume (m³ per hectare, top panel) and proportion in bottom panel.  Almost identical proportions were obtained when comparing basal area and tree density.")`
